---
title: "PS270 Final Project"
author: "William Klein"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to elections and polling, policies in different states especially their before and effects (with an emphasis on policies addressing housing crises and homelessness), canvassing and social engineering (nudge theory). This may change or be expanded upon depending on how available certain data is and my interests later in the semester.


## Proposal


Do homelessness rates across U.S. states increase with their cost of living? In this study, I plan to analyze the extent to which states cost-of-living indexes (COLI) affect their rates of homelessness. I hypothesize that if a state has a higher COLI, it will have a higher rate of homelessness per 10,000 people. In states with a higher cost of living, I suspect that the overall rate of homelessness will be higher. With higher cost-of-living indexes, basic necessities like food, housing, and utilities require a greater share of a person's finances. With this, it is only reasonable to assume that there would be more particular strain on low-income and vulnerable individuals, making it easier to fall into and harder to escape cycles of homelessness. My sample is comprised of data on homelessness and unemployment rates, cost-of-living indexes, along with the factors used to calculate it, and more, across all 50 states and the District of Columbia. My unit of analysis is State (e.g., Hawaii). The explanatory variable of interest is the cost-of-living index (COLI). That is how much a state is above or below the national average of cost of living, with a value of 100 being the national average. Any value over 100 is above the national average and any value below is lower than the national average. My outcome variable is homelessness, with it being the amount of homeless per 10k in a state. For example, Alabama has a homelessness value of 7.39, which means it has an average of 7.39 people experiencing homelessness per 10k. These variables are taken from Thomas F. Heston's data set on the cost-of-living and homelessness. If I observe a positive relationship between rates of homelessness and cost-of-living indexes in states, this would suggest my hypothesis to be true. As in, homelessness rates increasing with higher cost-of-living indexes and decreasing with lower. On the other hand, if I observe a negative relationship, such as higher rates of homelessness in states with lower cost-of-living indexes, this would prove my hypothesis to be false. When running my regression analysis on this, a positive and significant coefficient would suggest support for my hypothesis. 









```{r}
library(tidyverse)
library(readxl)
```

```{r}
Homeless <- read_xlsx("data/20230919-DATA-SOURCES.xlsx")
Homeless
```

```{r}
head(Homeless)
```

